\textbf{A framework for effective application of machine learning to microbiome-based classification problems}

### 2020-01-28 submission/manuscript.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={A framework for effective application of machine learning to a microbiome-based classification problem},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}
\usepackage[labelfont=bf]{caption}

@@

\title{\textbf{A framework for effective application of machine learning to a
microbiome-based classification problem}}
\author{}
\date{\vspace{-2.5em}}

@@

We also compared the top 20 OTUs selected by permutation importance in
L2-regularized logistic regression and the highest performing tree-based
models, random forest and XGBoost. 2 and 5 OTUs, respectively, were
shared among the models. These results indicate that we were able to
identify important OTUs that are shared across the highest performing
linear and non-linear models when we use permutation importance as our
interpretation method.

@@

\textbf{The computational efficiency of each ML model.} We compared the
training times of the seven ML models. The training times increased with
the complexity of the model and the number of potential hyperparameter
combinations. Also, the linear models trained faster than non-linear
models {[}Figures S1-S2; Figure 5{]}.

@@

\textbf{Figure 5. Training times of seven ML models.} The median
training time was the highest for XGBoost and shortest for
L2-regularized logistic regression. \newpage

\captionof{table}{\textbf{Table S1.} An aspirational rubric for evaluating the rigor of ML practices applied to microbiome data.}
\small
\begin{tabular}{|l|l|l|l|}
\hline

\rowcolor{lightgray}
\textbf{Practice} & \textbf{Poor} & \textbf{Good} & \textbf{Better} \\ \hline

\makecell[l]{Source \\ of data} & \makecell[l]{Data do not reflect intended \\  application (e.g., data pertain \\ to only patients with carcinomas \\ but model is expected to \\ predict advanced adenomas).} & \makecell[l]{Data are appropriate \\ for intended application.} & \makecell[l]{Data reflect intended \\ use and will persist \\ (e.g., same OTU assignments \\ for new fecal samples).} \\ \hline

\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population to which \\ the model will be applied.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline

\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method.} & \makecell[l]{Model choice is justified \\ for intended application.} & \makecell[l]{Different modeling choices \\ (justified for intended \\ application) are tested.} \\ \hline

\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning.} & \makecell[l]{Different hyperparameter \\ settings are explored \\ on training data.} & \makecell[l]{Hyperparameter grid search \\ performed by cross-validation \\ on the training set.} \\ \hline

\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the \\data used to train the model.} & \makecell[l]{Performance reported on \\ held-out test data.} & \makecell[l]{Performance reported on \\ multiple held-out test sets.} \\ \hline

\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported performance according to \\ a metric that is not appropriate \\ for intended application  (e.g., when \\  predicting rare outcome, accuracy \\ metric is not reliable).} & \makecell[l]{Reported performance in \\ terms of a metric that \\ is appropriate for intended \\ application and includes \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline

\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation.} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance.} & \makecell[l]{Hypotheses based on \\ feature importances \\ are generated and tested.} \\ \hline

\end{tabular}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={A framework for effective application of machine learning to microbiome-based classification problems},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}
\usepackage[labelfont=bf]{caption}

@@

\title{\textbf{A framework for effective application of machine learning to
microbiome-based classification problems}}
\author{}
\date{\vspace{-2.5em}}

@@

We also compared the top 20 OTUs selected by permutation importance in
L2-regularized logistic regression {[}Figure S7{]} and the highest
performing tree-based models, random forest and XGBoost {[}Figure 4{]}.
2 and 5 OTUs, respectively, were shared among the models. These results
indicate that we were able to identify important OTUs that are shared
across the highest performing linear and non-linear models when we use
permutation importance as our interpretation method.

@@

\textbf{The computational efficiency of each ML model.} We compared the
training times of the seven ML models. The training times increased with
the complexity of the model and the number of potential hyperparameter
combinations. Also, the linear models trained faster than non-linear
models {[}Figures S2-S3; Figure 5{]}.

@@

\hfill\break

\textbf{Figure 5. Training times of seven ML models.} The median
training time was the highest for XGBoost and shortest for
L2-regularized logistic regression.
